A Simple Type Theory for Category Theory

### 2022-02-28 paper/lipics-v2021-sample-article.tex
fibrations??

One final multiplicative we can define is the \emph{graph} of a
profunctor. This is unique among the multiplicatives in that it
defines a \emph{category} rather than a profunctor. It is a negative
multiplicative but it looks like a positive ??. A functor into the
category of elements $\exists (\alpha; \beta). R$ is equivalent to a
morphism into the functor $R$.

@@

Semantically, the graph of a profunctor is simply the category of
elements of the profunctor. Taking the profunctor to be (restrictions
of) the hom profunctor allows us to define the slice, arrow and comma
categories.

@@

\subsection{Adjoint Functors}

@@

Then we can show that any right-representation isomorphism is
determined by a co-unit:
\begin{lemma}[(Co)-unit of a Representation]
  If $G: \varr {\cat D}{\cat C}$ right-represents $R : \harr {\cat C}{\cat D}$, then there is a natural element

  \[ \pendallXdotY \beta \harrapp R {G\beta} \beta \]

  Such that the half of the isomorphism

  \[ \punitinXfromYtoZ {\cat C} {\alpha} {G \beta} \multimap \harrapp R {\alpha}{\beta} \]

  is defined by composition with the natural element.

  Dually, any $F$ left-representing isomorphism with $R$ is determined
  by a term of type $\pendallXdotY \alpha \harrapp R {\alpha}
  {F\alpha}$.
\end{lemma}
\begin{proof}
  A homomorphism $\punitinXfromYtoZ {\cat C} {\alpha} {G \beta} \multimap \harrapp R {\alpha}{\beta}$ is by definition a term of type
  \[ \pendallXdotY \alpha \homrallXYtoZ {\beta} {(\punitinXfromYtoZ {\cat C} \alpha {G \beta})} {\harrapp R \alpha \beta} \]
  By \cref{lem:fubini}, this type is isomorphic to 
  \[ \pendallXdotY \beta \homlallXYtoZ {\alpha} {(\punitinXfromYtoZ {\cat C} \alpha {G \beta})} {\harrapp R \alpha \beta} \]
  By \cref{lem:yo-and-co-yo}, this type is isomorphic to
  \[ \pendallXdotY \beta {{\harrapp R {(G \beta)} \beta}} \]

@@

>>>>

One final multiplicative we can define is the \emph{graph} of a
profunctor. Given a profunctor $\alpha:\cat C;\beta:\cat D \vdash R$,
the ``graph'' of $R$ is a category $\exists \alpha;\beta. R$. In
ordinary category theory, this is usually defined as the category of
elements of $R$ viewed as a functor $R : \cat C^o \times D \to \Set$.
In detail, its objects are triples $(a:\cat C, r : R, b : \cat D)$ and
morphism from $(a,r,b)$ to $(a', r', b')$ consists of a pair of
morphisms $g : b \to b'$ and a backwards $f : a' \to a$ such that $r'
= g\circ r \circ f$.
%
By picking $R = \punitinXfromYtoZ {\cat C} {\alpha}{\alpha'}$ we
recover the \emph{arrow} category $\exists
(\alpha;\alpha'). \punitinXfromYtoZ {\cat C}{\alpha}{\alpha'}$, by
composing these with functors, we recover the notion of \emph{comma}
category $\exists \alpha;\beta. \punitinXfromYtoZ {\cat
  E}{(F\alpha)}{(G\beta)}$ and by picking either $F$ or $G$ to have
domain the terminal category $1$, we recover the notion of a
\emph{slice} category.

This is unique among the multiplicatives in that it defines a
\emph{category} rather than a profunctor. Also note that while similar
to the tensor in that we use existential notation, the graph of a
profunctor is a \emph{negative} multiplicative, analogous to a
$\Sigma$-type. The universal property says that a functor from $\cat
E$ into the graph $\exists (\alpha; \beta). R$ is equivalent to a
transformation from $\cat E$ into the functor $R$, including functors
from $\cat E$ to $\cat C$ and $\cat D$.

@@

@@

\subsection{Adjoint Functors and Representable Profunctors}

@@

Then we can show that any right-representation isomorphism is
determined by a co-unit:
\begin{lemma}[(Co)-unit of a Representation]
  If $G: \varr {\cat D}{\cat C}$ right-represents $R : \harr {\cat
    C}{\cat D}$, then there is a natural element, the co-unit:
  \[ \pendallXdotY \beta \harrapp R {G\beta} \beta \]
  Such that the half of the isomorphism 
  \[ \punitinXfromYtoZ {\cat C} {\alpha} {G \beta} \multimap \harrapp R {\alpha}{\beta} \]
  is defined by composition with the natural element.

  Dually, any $F$ left-representing isomorphism with $R$ is determined
  by a ``unit'' of type $\pendallXdotY \alpha \harrapp R {\alpha}
  {F\alpha}$.
\end{lemma}
\begin{proof}
  A homomorphism $\punitinXfromYtoZ {\cat C} {\alpha} {G \beta} \multimap \harrapp R {\alpha}{\beta}$ is by definition a term of type
  \[ \pendallXdotY \alpha \homrallXYtoZ {\beta} {(\punitinXfromYtoZ {\cat C} \alpha {G \beta})} {\harrapp R \alpha \beta} \]
  By \cref{lem:fubini}, this type is isomorphic to 
  \[ \pendallXdotY \beta \homlallXYtoZ {\alpha} {(\punitinXfromYtoZ {\cat C} \alpha {G \beta})} {\harrapp R \alpha \beta} \]
  By \cref{lem:yo-and-co-yo}, this type is isomorphic to
  \[ \pendallXdotY \beta {{\harrapp R {(G \beta)} \beta}} \]

@@

\subsection{(op)-Fibrations}

Using the graph of a profunctor, we can give the following definition
for when a functor is a fibration. This gives a very weak notion of
fibration that is probably equivalent to a Street fibration (todo: is
this an existing definition???).

\begin{definition}{Street Fibration}
  Let $F : \varr{\cat C}{\cat D}$, we say $F$ is a \emph{fibration}
  when the induced functor
  \[p : \exists {\alpha:\cat C;\alpha':\cat C} \punitinXfromYtoZ {\cat C}{\alpha}{\alpha'} \vdash (F\pi_-p,F_1\pi_pp, \pi_+p) : \exists {\beta:\cat D;\alpha':\cat C}. \punitinXfromYtoZ {\cat D} \beta {F \alpha'} \]
  has a right adjoint inverse.

  Dually, $F$ is an \emph{op-fibration} when the same induced functor
  has a left adjoint inverse.
\end{definition}

We can get a strict notion of fibration if we allow ourselves to be a little ``evil'':

\begin{definition}{Grothendieck Fibration}
  Let $F : \varr{\cat C}{\cat D}$, we say $F$ is a \emph{strict
  fibration} when the induced functor $F'$ has a strict right adjoint
  inverse $R$, i.e. inverse $\lambda p. R(F'(p)) = \lambda p. p$. We
  analogously define a strict op-fibration.  
\end{definition}

Alternatively, we can give a more traditional definition using slice
categories, but the downside is the formulation is inevitably higher
order.
\begin{definition}{Street Fibration}
  Let $F : \varr{\cat C}{\cat D}$, we say $F$ is a \emph{fibration}
  when for every $a : \varr{1}{\cat C}$, the induced functor
  \[p : \exists {\alpha:\cat C; o:\cat 1} \punitinXfromYtoZ {\cat C}{\alpha}{a()} \vdash (F\pi_-p,F_1\pi_pp, \pi_+p) : \exists {\beta:\cat D; o:\cat 1}. \punitinXfromYtoZ {\cat D} \beta {F(a())} \]
  has a right adjoint inverse.

  Dually, $F$ is an \emph{op-fibration} when the same induced functor
  has a left adjoint inverse.
\end{definition}

\begin{lemma}{Fibrations compose}
  If $F : \varr{\cat C}{\cat D}$ and $G : \varr{\cat D}{\cat E}$ are
  (op)-fibrations then so is $G \circ F$.
\end{lemma}
\begin{proof}
  By composition of adjoints.
\end{proof}
